### 2019-10-13 ch1_introduction/content.tex
update intro

\subsection{Network inference}
%One of the central cellular processes underlying development is transcriptional regulation. During development, changes in transcription factor activity induce chromatin modifications, chromatin remodeling and ultimately a differential recruitment of the basal transcriptional machinery [66]. Modeling the dynamics of gene regulation is therefore essential to better understand why a cellular dynamic processes progresses through several steps, and what goes wrong in the case of disease.
%
%The dynamics of gene regulation has classically been studied using time series data [23]. When dynamic processes progress asynchronously, such as in hematopoiesis, time series data are usually obtained by sorting different transition states and assessing bulk gene expression and transcription factor binding within the population [67–70]. Alternatively, time series data can also be generated by synchronizing the dynamic process between cells. However, issues with time-resolution, heterogeneity and good in vivo synchronization models can often limit the predictive power of the dynamic models of gene regulation which can be constructed [23]. Combining single-cell snapshot data with trajectory inference can now in principle construct an accurate time-series dataset for every cell population both in vivo and in vitro, while avoiding issues with heterogeneity. Initially, several pioneering studies demonstrated the potential of this approach in single cells using single-cell qPCR data for selected sets of 10 to 50 transcription factors [71–75]. With the advent of genome wide single-cell technologies, the models of dynamic gene regulation can now be made even more complete while avoiding a prior bias to well studied transcription factors [17,30,76,77] and several methods have already been developed specifically tuned towards single-cell data [78–80].
%
%The power of a dynamic network approach in single cells is not only that it allows the identification of transcription factors specifically regulating a specific state, but also how these factors then induce sets of genes necessary for regulating the next wave of transcription changes (Figure 1B). Such a gene regulatory cascade was described during cellular reprogramming from mouse embryonic fibroblasts to induced pluripotent stem cells, where SOX2 was shown to activate a series of other transcription factors which, together with SOX2, induces pluripotency [72]. Transition states are frequently associated with particular subnetworks of transcriptional regulators, as was shown in a direct reprogramming setting from fibroblasts to neurons [77]. Several studies have highlighted how some regulatory links can be very dynamic while others show evidence of being static during consecutive developmental stages [74,75]. Regulatory network motifs can also be extracted from the regulatory network [74], some of which can have interesting properties in dynamic systems [81,82]. In branching trajectories dynamic regulatory networks can also be used to study the regulators important for a particular branch [17,75,79] (Figure 1C).
%
%Another advantage of using ordered datasets, either based on bulk time-series or single-cell trajectories, is the improved accuracy of inferring true causative regulatory interactions. Assume that during development the expression of regulator A increases and in turn this regulator activates the expression of regulator B and gene C. Between the upregulation of regulator A and the upregulation of B and C we expect a time delay, equal to the time it takes for regulator A to be translated, translocated to the nucleus and induce a detectable increase in the expression of B and C. On the other hand the expression of regulator B and gene C will be very strongly correlated, because they are both activated at the same time. Incorporating this time-delay into the gene regulatory network inference could therefore be used to better discern the true causative regulatory edge between A and B/C with a spurious correlative edge between B and C [83].

Gene regulatory network inference, or network inference (NI) for short, is a type of computational analysis where thousands of transcriptomic profiles are analysed together in order to infer the regulatory interactions between transcription factors and genes. This topic already received much attention with the advent of bulk omics (before single-cell omics). These efforts culminated in several DREAM competitions assessing the performance of 29 different NI methods\cite{marbach_revealingstrengthsweaknesses_2010, marbach_wisdomcrowdsrobust_2012}.

After the last DREAM competition, it seemed that interest in NI methodology had declined. After all, NI on bulk omics profiles suffered from several crucial issues. As mentioned previously, bulk profiles are generated by pooling together the RNA transcripts of a supposedly homogeneous population of thousands of cells. Since the expression values are averaged over the whole population, incorrect assumptions on the homogeneity of the pooled cells may lead to the masking of relevant expression patterns in rare cell populations (Figure \ref{fig:bulk_vs_singlecell}). Besides, NI methods rely on a diverse set of time-series and perturbation experiments in order to reliably identify causal regulatory interactions. Such experiments are expensive and time-consuming, and an inaccurate selection of time points might result in crucial intermediate stages being missed.

\begin{figure}[htb!]
	\centering
	\includegraphics[width=\Largefigure]{fig/ni_bulk_vs_sc}
	\caption{Bulk expression data return the average expressions of genes among large numbers of cells. In order to infer regulatory networks from this type of data, multiple bulk profiles (resulting from time series or perturbation experiments) are required. On the other hand, sequencing the transcriptome at the single-cell level uncovers the high variability among cells, providing the necessary information to infer gene regulatory networks directly.} % TODO: Already used this figure in the previous section
	\label{fig:bulk_vs_singlecell}
\end{figure}

The advent of single-cell omics has made scientists wonder whether now is the time to revisit network inference\cite{stegle_computationalanalyticalchallenges_2015}. 
One of the main advantages of single-cell omics is the ability to quantify the exact cellular state of thousands of cells per experiment. The heterogeneity between cells caused by naturally occurring biological randomness\cite{padovan-merhar_usingvariabilitygene_2013} can be exploited to infer regulatory interactions between TFs and their target genes at much lower costs (see Figure \ref{fig:bulk_vs_singlecell}).
In this setting, heterogeneity in the cell population eases network inference, rather than mask condition-specific expression patterns in regulatory interactions.

% mention differential networks

\subsection{RNA velocity}

\subsection{Trajectory alignment}

@@

\begin{itemize}
	\item We develop benchmarking strategies for assessing the performance of computational tools constrained by low availability of novel types of real single-cell data (Chapter~\ref{chap:dyngen}). \textit{In silico} simulations of individual cells are used to help kick-start emerging domains much more safely and allow anticipation of future technological developments by already developing computational tools.
	\item We apply this strategy to perform a comparison of TI methods (Chapter~\ref{chap:dynbenchmark}). Trajectory inference is one of the largest categories of all the novel single-cell omics tools, yet a comprehensive and quantitative study of the advantages and disadvantages of the numerous tools was hitherto lacking. We provide a set of guidelines for end-users wishing to infer trajectories. We also make our pipeline, datasets, metrics, and containerised wrappers of TI methods publicly available for developers to use.
	\item We developed dyno, a toolkit to easily infer, visualise and interpret single-cell trajectories using more than 50 different TI methods (Chapter~\ref{chap:dyno}). dyno provides downstream analysis such as: visualising a trajectory in a low-dimensional space or a heatmap, detecting genes differentially expressed at different stages of the tragectory, comparing multiple trajectories in a common dimensionality reduction, and manipulating the trajectory (e.g. adding directionality or adding annotation).
	\item We introduce a novel TI method specialised in inferring linear trajectories (Chapter~\ref{chap:scorpius}). Despite linear TI being the most simple but commonly used form of trajectory inference, the benchmark demonstrated that most TI methods are not capable of producing accurate models of linear datasets.
	\item We invent a new type of NI method capable of inferring the GRN of individual cells (Chapter~\ref{chap:bred}). We demonstrate this <yadeyade .. fill in when the chapter is actually written.>
	\item Every NI method has certain topological biases. We provide a tool for analysing the topological properties of large, evolving networks and use this to iteratively optimise GRN predictions (Chapter~\ref{chap:incgraph}).
	\item We discuss reproducibility problems of TI methods due to low rates of quantitative self-assessment (Chapter~\ref{chap:selfassessment}). We provide solutions for different causal reasons for this phenomenon in order to spur developers to perform more self-assessments.
	\item Finally, we summarise our experience in benchmarking computational methods in a list of essential guidelines (Chapter~\ref{chap:guidelines}).
\end{itemize}

>>>>

\subsection{Network inference}
One of the central cellular processes underlying development is transcriptional regulation. 	
Modeling the dynamics of gene regulation is therefore essential to better understand why a cellular dynamic processes progresses through several steps, and what goes wrong in the case of disease.

Network inference (NI) methods predict which genes are regulated by which transcription factors (Figure~\ref{fig:comp_tools_ni}). Before single-cell omics, these methods rely on multiple experiments, amongst which perturbation and time-series experiments, to predict the effect each transcription factor has on the up- or downregulation of a gene.

One of the main advantages of single-cell omics is the heterogeneity between cells caused by naturally occurring biological randomness\cite{padovan-merhar_usingvariabilitygene_2013} can be exploited to infer regulatory interactions between TFs and their target genes at much lower costs.

\begin{figure}[htb!]
	\centering
	\includegraphics[width=.6\linewidth]{fig/comptools2/comptool_5_networkinference.pdf}
	\caption{
		\textbf{Network inference for single-cell omics data.}
	}
	\label{fig:comp_tools_ni}
\end{figure}

@@

\begin{itemize}
	\item We develop benchmarking strategies for assessing the performance of computational tools constrained by low availability of novel types of real single-cell data (Chapter~\ref{chap:dyngen}). \textit{In silico} simulations of individual cells are used to help kick-start emerging domains much more safely and allow anticipation of future technological developments by already developing computational tools.
	\item We apply this strategy to perform a comparison of TI methods (Chapter~\ref{chap:dynbenchmark}). Trajectory inference is one of the largest categories of all the novel single-cell omics tools, yet a comprehensive and quantitative study of the advantages and disadvantages of the numerous tools was hitherto lacking. We provide a set of guidelines for end-users wishing to infer trajectories. We also make our pipeline, datasets, metrics, and containerised wrappers of TI methods publicly available for developers to use.
	\item We developed dyno, a toolkit to easily infer, visualise and interpret single-cell trajectories using more than 50 different TI methods (Chapter~\ref{chap:dyno}). dyno provides downstream analysis such as: visualising a trajectory in a low-dimensional space or a heatmap, detecting genes differentially expressed at different stages of the tragectory, comparing multiple trajectories in a common dimensionality reduction, and manipulating the trajectory (e.g. adding directionality or adding annotation).
	\item We introduce a novel TI method specialised in inferring linear trajectories (Chapter~\ref{chap:scorpius}). Despite linear TI being the most simple but commonly used form of trajectory inference, the benchmark demonstrated that most TI methods are not capable of producing accurate models of linear datasets.
	\item We invent a new type of NI method capable of inferring the GRN of individual cells (Chapter~\ref{chap:bred}). We demonstrate this <yadeyade .. fill in when the chapter is actually written.>
	% TODO: fill this part in!!!
	\item Every NI method has certain topological biases. We provide a tool for analysing the topological properties of large, evolving networks and use this to iteratively optimise GRN predictions (Chapter~\ref{chap:incgraph}).
	\item We discuss reproducibility problems of TI methods due to low rates of quantitative self-assessment (Chapter~\ref{chap:selfassessment}). We provide solutions for different causal reasons for this phenomenon in order to spur developers to perform more self-assessments.
	\item Finally, we summarise our experience in benchmarking computational methods in a list of essential guidelines (Chapter~\ref{chap:guidelines}).
\end{itemize}
